Pillar 1 - Student Topic Profile

### 2019-06-27 Conference/COPA2019/paper/conference figures COPA2019.Rmd
no message

```{r student_data}
set.seed(2019)

@@

  arrange(`Academic Year`) %>%

  print
```

```{r course_data}
course_data <- d_text$catalogues %>%

@@

  select(`Course ID`, `Course Title`, Department = department, word) %>%

  print
```

```{r fig_student_data}
student_data %>% latex_table("data-student")
```

```{r fig_course_data}
course_data %>% latex_table("course-student")
```

@@

@@

@@

```{r}
d_width <- conformal_results %>%
  group_by(course, significance) %>%
  summarise(q_50 = quantile(width, 0.5), # median
            q_10 = quantile(width, 0.1),
            q_90 = quantile(width, 0.9))
```

```{r}
d_width %>%
  ggplot() +
  geom_errorbar(aes(significance, ymax = q_90, ymin = q_10), size = .5) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_line(aes(significance, q_50, group = 1)) +
  geom_point(aes(significance, q_50)) +
  facet_wrap( ~ course) +
  labs(y = "Width",
       x = "Significance Level") +
  scale_x_continuous(breaks = c(0.05, 0.2, 0.4, 0.6, 0.8))

@@

```{r}
d_error <- conformal_results %>%
  group_by(course, significance) %>%
  summarise(error = mean(!hit))
```

```{r}
d_error %>%
  ggplot(aes(x = significance, y = error)) +
  geom_abline(slope = 1, linetype = "longdash") +
  geom_line(aes(group = 1)) +
  geom_point() +
  facet_wrap( ~ course) +
  labs(y = "Error Rate",
       x = "Significance Level") +
  scale_x_continuous(breaks = c(0.05, 0.2, 0.4, 0.6, 0.8))

## Table
```{r}
conformal_results %>%
  filter(significance < 0.2) %>%
  group_by(course, significance) %>%
  summarize(width_med  = median(width),
            error_rate = mean(!hit)) %>%

  gather(key = "measure", value = "output", width_med, error_rate) %>%
  mutate(output = round(output, 2)) %>%
  unite(id, measure, significance, sep = "-") %>%
  spread(key = id, value = output) %>%

>>>>

```{r fig_student_data}
set.seed(2019)

@@

  arrange(`Academic Year`)

student_data %>% latex_table("data-student")
```

```{r fig_course_data}
course_data <- d_text$catalogues %>%

@@

  select(`Course ID`, `Course Title`, Department = department, word)

course_data %>% latex_table("course-student")
```

@@

```{r}
plot_width <- 8
plot_height <- plot_width / 1.6 
```

@@

```{r}
d_error <- conformal_results %>%
  group_by(course, significance, underl_algo) %>%
  summarise(error = mean(!hit))
```

```{r}
d_error %>%
  ggplot(aes(x = significance, y = error, col = underl_algo)) +
  geom_abline(slope = 1, linetype = "longdash") +
  geom_line(aes(group = underl_algo)) +
  geom_point() +
  facet_wrap( ~ course) +
  labs(y = "Error Rate", x = "Significance Level") +
  scale_x_continuous(breaks = c(0.05, 0.2, 0.4, 0.6, 0.8)) +
  scale_color_discrete(name = "Underlying Algorithm", labels = c("Boosting", "Lasso","Random Forest"))

@@

```{r}
d_width <- conformal_results %>%
  group_by(course, significance, underl_algo) %>%
  summarise(q_50 = quantile(width_corrected, 0.5), # median
            q_10 = quantile(width_corrected, 0.1),
            q_90 = quantile(width_corrected, 0.9))
```

```{r}
dodge_width <- 0.04

d_width %>%
  ggplot(aes(col = underl_algo)) +
  geom_errorbar(aes(significance, ymax = q_90, ymin = q_10), size = .5, position = position_dodge(width = dodge_width)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_line(aes(significance, q_50, col = underl_algo), position = position_dodge(width = dodge_width), linetype = 2) +
  geom_point(aes(significance, q_50), position = position_dodge(width = dodge_width)) +
  facet_wrap( ~ course) +
  labs(y = "Width", x = "Significance Level") +
  scale_x_continuous(breaks = c(0.05, 0.2, 0.4, 0.6, 0.8)) +
  scale_color_discrete(name = "Underlying Algorithm", labels = c("Boosting", "Lasso","Random Forest"))

@@

```{r}
conformal_results %>%
  ggplot() +
  geom_boxplot(aes(significance, width_corrected, group = significance, col = underl_algo)) +
  facet_wrap(~course)
```

## Table
```{r}
conformal_results %>%
  filter(significance <= 0.2) %>%
  group_by(course, significance, underl_algo) %>%
  summarize(width_med  = median(width_corrected),
            error_rate = mean(!hit)) %>%
  gather(key = "measure", value = "output", width_med, error_rate) %>%
  mutate(output = round(output, 2)) %>%
#  unite(col = "ID", underl_algo, measure, sep = " ") %>%
  unite(col = ID, measure, significance, underl_algo, sep = "-") %>%
  spread(key = ID, value = output) %>%

### 2019-06-27 Data Preparation - Student Data.Rmd
no message

>>>>

## Censored data for Research Project

```{r}
student_ID <- d_transcript_augmented %>% distinct(`Student ID`)
randomID <- sample(1 : nrow(student_ID))
student_ID_random <- student_ID %>% mutate(Student_ID_random = randomID)

d_transcript_censored <- d_transcript_augmented %>%
  left_join(student_ID_random, by = "Student ID") %>%
  select(Student_ID_random, Course_ID = `Course ID`, time, Concentration, Cluster,
         Academic_year = `Academic Year`, Period = period_numerical) %>%
  sample_frac(1, replace = FALSE)

d_transcript_censored

write_csv(d_transcript_censored, path = "Output/d_transcript_censored.csv")
```

### 2019-06-27 Warning Issuance - Conformal Prediction.Rmd
no message

```{r setup, include = FALSE}
set.seed(1)

@@

library(tidyverse)
library(glmnet)

@@

  student_profile %>%
    filter(`Course ID` == course) %>%
    select(Y = Grade, matches("GPA|Topic"))

@@

## Fit lasso on training set
```{r,lasso helper}
fit_target_model <- function(df){

  X <- df %>% select(matches("GPA|Topic"))
  Y <- df %>% pull(Y)

  cv.glmnet(x = as.matrix(X), y = Y, type.measure = "mae") # lasso by default. Use mean absolute error

@@

predict_grade <- function(df, model){

  X <- df %>% select(matches("GPA|Topic"))

  Y_hat <- predict.cv.glmnet(model, newx = as.matrix(X), s = "lambda.min") # use lambda that minimizes mae

@@

fit_error_model <- function(df){

  X  <- df %>% select(matches("GPA|Topic"))
  mu <- df %>% pull(mu)

  cv.glmnet(x = as.matrix(X), y = mu, type.measure = "mae")

@@

predict_error <- function(df, model){

  X <- df %>% select(matches("GPA|Topic"))

  mu_hat <- predict.cv.glmnet(model, newx = as.matrix(X), s = "lambda.min")

@@

## Compute standardized non-conformity scores on callibration test
```{r}

compute_ncs              <- function(df) df %>% mutate(alpha = abs(Y - Y_hat))

compute_standardized_ncs <- function(df) df %>% mutate(sigma = exp(mu_hat),
                                                       alpha = abs(Y - Y_hat) / sigma )

@@

  max(df$alpha) + 0.00001

@@

  df %>% mutate(
    sigma       = exp(mu_hat),
    width       = ncs * sigma,
    border_low  = Y_hat - width,
    border_high = Y_hat + width
    )
}

compute_hit <- function(df){
  df %>% mutate(hit = (border_low <= Y) & (Y <= border_high))
}
```

@@

significance <- c(0.01, 0.05, seq(0.1, 0.9, 0.1))
results <- list()
n_list <- 1

# course <- "SSC3044"; iteration <- 1; signi <- 0.1

@@

@@

    # fit target and error model to proper training set
    model_target      <- fit_target_model(Z_proper_training)
    Z_proper_training <- predict_grade(Z_proper_training, model_target)
    model_error       <- fit_error_model(Z_proper_training)

    # non-conformity scores on calibration
    Z_calibration <- Z_calibration %>% 
      predict_grade(model_target) %>%
      predict_error(model_error) %>%
      compute_standardized_ncs %>%
      compute_significance

    # error on test set
    Z_test <- Z_test %>% 
      predict_grade(model_target) %>%
      predict_error(model_error)

@@

      Z_test_augmented <- Z_test %>% 
        compute_interval(ncs = ncs) %>%
        compute_hit

      # save results
      results[[n_list]] <- Z_test_augmented %>%
        mutate(significance = signi,
               iteration    = iteration,
               course       = course) %>%
        select(course, iteration, significance, width, hit,
               Y, Y_hat, error, mu, mu_hat, sigma)

@@

  } # end CV fold

```{r}
conformal_results <- bind_rows(results) %>%
    mutate(iteration    = as.factor(iteration),
           course       = as.factor(course))

@@

Illustration of the effect of error_hat on interval's width (for significance of 99%):
```{r, fig.height=4, eval = F}
Z_test %>% select(Y, Y_hat, mu_hat, width, border_low, border_high, hit)
plot(Z_test$error, Z_test$mu_hat)
```

Correlation between the actual errors and the predicted errors (error_hat):
```{r, eval = F}
cor(Z_test$error, Z_test$error_hat)
```

```{r, echo = FALSE, eval = F}
ggplot(data = Z_callibration) + geom_histogram(aes(Y), bins = 15)

ggplot(data = Z_callibration) + geom_histogram(aes(Y_hat), bins = 15)

ggplot(data = Z_callibration) + geom_histogram(aes(alpha), bins = 15)
```

```{r, echo = FALSE, eval = F}
ggplot(data = Z_callibration) + geom_point(aes(Y, Y_hat))

ggplot(data = Z_callibration) + geom_point(aes(Y, alpha))

ggplot(data = Z_callibration) + geom_point(aes(Y_hat, alpha))
```

>>>>

```{r setup, include = FALSE}
time0 <- Sys.time()

set.seed(1)

@@

library(tidyverse)
library(glmnet)
library(randomForest)
library(xgboost)

@@

  student_model %>%
    filter(`Course ID` == course) %>%
    select(Y = Grade, matches("GPA|Topic"))

@@

## Fit lasso on training set
```{r,lasso helper}
fit_target_model <- function(df, underlying_algorithm = underlying_algo){

  X <- df %>% select(matches("GPA|Topic")) %>% as.matrix
  Y <- df %>% pull(Y)

  if(underlying_algorithm == "lasso"){
    cv.glmnet(x = X, y = Y,
              type.measure = "mae") # lasso by default. Use mean absolute error

  }else if(underlying_algorithm == "randomforest"){
    randomForest(x = X, y = Y,
                 ntree = 1000,
                 mtry = floor(ncol(X)/3),
                 nodesize = 5)

  }else if(underlying_algorithm == "boost"){
    xgboost(data = X, label = Y,
            max_depth = 2, eta = .3, nrounds = 1000,
            objective = "reg:linear", verbose = 0)

  }else{stop("Unknown algorithm")}

@@

predict_grade <- function(df, model, underlying_algorithm = underlying_algo){

  X <- df %>% select(matches("GPA|Topic")) %>% as.matrix

  if(underlying_algorithm == "lasso"){
    Y_hat <- predict.cv.glmnet(model, newx = X, s = "lambda.min") # use lambda that minimizes mae

  }else if(underlying_algorithm == "randomforest"){
    Y_hat <- predict(model, X)

  }else if(underlying_algorithm == "boost"){
    Y_hat <- predict(model, X)

  }else{stop("Unknown algorithm")}

@@

fit_error_model <- function(df, underlying_algorithm = underlying_algo){

  X  <- df %>% select(matches("GPA|Topic")) %>% as.matrix
  mu <- df %>% pull(mu)

  if(underlying_algorithm == "lasso"){
    cv.glmnet(x = X, y = mu, type.measure = "mae")

  }else if(underlying_algorithm == "randomforest"){
    randomForest(x = X, y = mu,
                 ntree = 1000,
                 mtry = floor(ncol(X)/3),
                 nodesize = 5)

  }else if(underlying_algorithm == "boost"){
    xgboost(data = X, label = mu,
            max_depth = 2, eta = .3, nrounds = 1000,
            objective = "reg:linear", verbose = 0)

  }else{stop("Unknown algorithm")}

@@

predict_error <- function(df, model, underlying_algorithm = underlying_algo){

  X <- df %>% select(matches("GPA|Topic")) %>% as.matrix

  if(underlying_algorithm == "lasso"){
    mu_hat <- predict.cv.glmnet(model, newx = X, s = "lambda.min")

  }else if(underlying_algorithm == "randomforest"){
    mu_hat <- predict(model, X)

  }else if(underlying_algorithm == "boost"){
    mu_hat <- predict(model, X)

  }else{stop("Unknown algorithm")}

@@

## Compute standardized non-conformity scores on callibration test
```{r}
compute_ncs              <- function(df) df %>% mutate(alpha = abs(Y - Y_hat))

compute_standardized_ncs <- function(df) df %>% mutate(sigma = exp(mu_hat),
                                                       sigma_corrected = sigma + gamma,
                                                       alpha = abs(Y - Y_hat) / sigma_corrected )

@@

  max(df$alpha) + 1e-5

@@

  df %>% mutate(
    sigma = exp(mu_hat),
    sigma_corrected = sigma + gamma,
    width           = ncs * sigma_corrected,
    border_low      = Y_hat - width,
    border_high     = Y_hat + width
    )
}

correct_interval <- function(df){

  df %>% mutate( # constraint prediction interval to [0, 10] 
    border_low_corrected  = pmax(0 , border_low),
    border_high_corrected = pmin(10, border_high),
    width_corrected       = border_high_corrected - border_low_corrected)

compute_hit <- function(df){
  df %>% mutate(hit = (border_low_corrected <= Y) & (Y <= border_high_corrected))
}
```

@@

set.seed(1)
significance <- c(0.05, seq(0.1, 0.9, 0.1))
results <- list()
n_list <- 1
underlying_algos <- c("boost", "lasso", "randomforest")
gamma <- 2

# course <- "SSC3044"; iteration <- 1; signi <- 0.1; underlying_algo =

@@

  for(underlying_algo in underlying_algos){

@@

      # fit target and error model on proper training set
      model_target      <- fit_target_model(Z_proper_training)
      Z_proper_training <- predict_grade(Z_proper_training, model_target)
      model_error       <- fit_error_model(Z_proper_training)

      # non-conformity scores on calibration set
      Z_calibration <- Z_calibration %>% 
        predict_grade(model_target) %>%
        predict_error(model_error) %>%
        compute_standardized_ncs %>%
        compute_significance

      # predict grade and error on test set
      Z_test <- Z_test %>% 
        predict_grade(model_target) %>%
        predict_error(model_error)

@@

        Z_test_augmented <- Z_test %>% 
          compute_interval(ncs = ncs) %>%
          correct_interval %>%
          compute_hit

        # save results
        results[[n_list]] <- Z_test_augmented %>%
          mutate(significance = signi,
                 iteration    = iteration,
                 course       = course,
                 underl_algo  = underlying_algo) %>%
          select(course, iteration, significance, width, hit,
                 Y, Y_hat, error, mu, mu_hat, sigma,
                 border_low, border_low_corrected, border_high, border_high_corrected, width_corrected,
                 underl_algo)

@@

    } # end loop CV fold

  } # end loop underlying_algorithms

} # end loop courses
```

```{r}
conformal_results <- bind_rows(results) %>%
    mutate(iteration    = as.factor(iteration),
           course       = as.factor(course),
           underl_algo  = as.factor(underl_algo))

@@

Illustration of the effect of error_hat on interval's width (for significance of 99%):
```{r, fig.height=4, eval = F}
conformal_results %>%
  ggplot() +
  geom_density(aes(Y)) +
  facet_wrap(~ course)

conformal_results %>%
  ggplot() +
  geom_density(aes(Y_hat, col = underl_algo)) +
  facet_wrap( ~ course)

conformal_results %>%
  ggplot() +
  geom_point(aes(Y, Y_hat), size = 0.1) +
  facet_grid(course ~ underl_algo)
```

```{r}
conformal_results %>%
  ggplot() +
  geom_point(aes(mu, mu_hat)) +
  facet_wrap(~ course)
```

Correlation between the actual errors and the predicted errors (error_hat):
```{r, eval = F}
conformal_results %>%
  group_by(course, underl_algo) %>%
  summarize(var_Y = var(Y),
            cor_Y = cor(Y, Y_hat),
            error = cor(mu, mu_hat))
```

# Timing
```{r}
Sys.time() - time0
```

### 2019-06-27 Warning Issuance - Grade Prediction with RF.Rmd
no message

# Random forest

```{r}
data(iris)

iris.rf <- randomForest(iris[,-5], iris[,5], prox=TRUE)

iris.p <- classCenter(iris[,-5], iris[,5], iris.rf$prox)

plot(iris[,3], iris[,4], pch=21, xlab=names(iris)[3], ylab=names(iris)[4],
bg=c("red", "blue", "green")[as.numeric(factor(iris$Species))],
main="Iris Data with Prototypes")

points(iris.p[,3], iris.p[,4], pch=21, cex=2, bg=c("red", "blue", "green"))
```

```{r}
rf1 <- randomForest(Species ~ ., iris, ntree=50, norm.votes=FALSE)

rf2 <- randomForest(Species ~ ., iris, ntree=50, norm.votes=FALSE)

rf3 <- randomForest(Species ~ ., iris, ntree=50, norm.votes=FALSE)

rf.all <- combine(rf1, rf2, rf3)
print(rf.all)
```

```{r}
## Look at the third trees in the forest.
getTree(randomForest(iris[,-5], iris[,5], ntree=10), 3, labelVar=TRUE)
```

```{r}
iris.rf <- randomForest(Species ~ ., iris, ntree=50, norm.votes=FALSE)
iris.rf <- grow(iris.rf, 50)
print(iris.rf)
```

```{r}
set.seed(4543)
data(mtcars)
mtcars.rf <- randomForest(mpg ~ ., data=mtcars, ntree=1000,
keep.forest=FALSE, importance=TRUE)

importance(mtcars.rf)
importance(mtcars.rf, type=1)
```

```{r}
data(imports85)
imp85 <- imports85[,-2] # Too many NAs in normalizedLosses.
imp85 <- imp85[complete.cases(imp85), ]
## Drop empty levels for factors.
imp85[] <- lapply(imp85, function(x) if (is.factor(x)) x[, drop=TRUE] else x)
stopifnot(require(randomForest))
price.rf <- randomForest(price ~ ., imp85, do.trace=10, ntree=100)
print(price.rf)
numDoors.rf <- randomForest(numOfDoors ~ ., imp85, do.trace=10, ntree=100)
print(numDoors.rf)
```

>>>>

# Extract sample of one course

```{r}
{z <- find_df("HUM2025") %>%
  select(Grade, matches("GPA|Topic"))

n_sample <- nrow(z)
n_train  <- floor(0.67 * n_sample)
n_test   <- n_sample - n_train

allocation <- sample(rep(c("train", "test"), times = c(n_train, n_test)))
z <- z %>% mutate(allocation = allocation)
z_train <- z %>% filter(allocation == "train") %>% select(-allocation)
z_test  <- z %>% filter(allocation == "test" ) %>% select(-allocation)} # prepare sample
```

# Grade prediction
```{r}
assess_prediction <- function(test_set = z_test, predictions){

  test_set <- test_set %>% mutate(y_hat = predictions,
                                  res = Grade - y_hat)

 # hist(test_set$y_hat)
#  plot(test_set$y_hat, test_set$res)
#  plot(test_set$Grade, test_set$res)
  plot(test_set$Grade, test_set$y_hat)

  mae <- test_set$res %>% abs %>% mean
  print(str_c("MAE: ", mae %>% round(2)))

}
```

```{r}
# data
x <- z_train %>% select(matches("Topic|GPA")) %>% as.matrix
y <- z_train %>% pull(Grade)
x_test <- z_test %>% select(matches("Topic|GPA"))

# Boost
m_boost <- xgboost(data = x, label = y,
                   max_depth = 2, eta = .3, subsample = 1, nrounds = 1000, colsample_bytree = .5,
                   objective = "reg:linear")
y_boost <- predict(m_boost, as.matrix(x_test))
assess_prediction(predictions = y_boost)

# RF
m_RF <- randomForest(as.matrix(x), y,
                     ntree = 1000, mtry = floor(ncol(z_train)/3))

y_rf <- predict(m_RF, as.matrix(x_test))
assess_prediction(predictions = y_rf)

m_lasso <- cv.glmnet(as.matrix(x), y,
                     nfolds = 10, type.measure = "mae")
y_lasso <- predict(m_lasso, newx = as.matrix(x_test), s="lambda.min")
assess_prediction(predictions = y_lasso)
```
